---
title: "Instacart - Data Understanding Report"
author: "Priyanka Ghule & Saurabh Chakrabarty"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)
```

## Exploratory Data Analysis

```{r libraries}
library(readr)          # To read data
# library(tidyverse)    # includes readr
library(data.table)     # For data manipulation
library(ggplot2)        # For plotting
library(gridExtra)      # To arrange ggplots
library(scales)         # To adjust numeric scales in ggplots
library(dplyr)          # To use pipes?
```

```{r load_data}
orders <- read_csv('../input/orders.csv')
orders <- as.data.table(orders)
```

### Days Since Prior Order

The below is a distribution of the days since prior order.

Some observations:

- There are a number of people who make orders on the same day (0 days)
- There are lots of people who make orders on a weekly basis (7 days)
- There are some people who make orders every fortnight (14 days)
- There are yet some people who make orders every 3 weeks (21 days)
- 30 days appears to represent either:
  + People making orders monthly
  + Orders made 30 or more days apart

```{r}
ggplot(orders, aes(x = days_since_prior_order)) +
  geom_bar(fill = c("pink", rep("grey25", 6), 
                    "pink", rep("grey25", 6), 
                    "pink", rep("grey25", 6),
                    "pink", rep("grey25", 6),
                    "pink", "grey50", "pink")) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(x = "Days Since Prior Order",
       y = "Count",
       title = "Distribution of Orders by Days Since Prior Order") +
  scale_y_continuous(labels = comma)
```

### Days Since Prior Order & Day of Week

Splitting the order data by day of week presents deviations from the overall pattern
shown above.

We can now start to see sub-populations in the data.

It appears that people who order on days 0 and 1 have  a tendency  to order 
7 days apart. This shows a population with a habbit of purchasing on the same 
day each week.

This pattern is not so pronounced  during days 2 - 4, though there is again for 
days 5 and 6. 

```{r}
## Facet using grid.arrange for greater control of layout and colours
po0 <- ggplot(orders[order_dow == 0, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 7), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 6), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 2))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 0")

po1 <- ggplot(orders[order_dow == 1, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 7), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 6), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 2))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 1")

po2 <- ggplot(orders[order_dow == 2, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 7), "orange",  rep("grey25", 6), "orange", "orange",
                    rep("grey25", 6), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 1))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 2")

po3 <- ggplot(orders[order_dow == 3, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 2), "orange",  rep("grey25", 28))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 3")

po4 <- ggplot(orders[order_dow == 4, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 3), "orange",  rep("grey25", 3), "orange",
                    rep("grey25", 23))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 4")

po5 <- ggplot(orders[order_dow == 5, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 4), "orange",  rep("grey25", 2), "orange",
                    rep("grey25", 23))) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 5")

po6 <- ggplot(orders[order_dow == 6, ], aes(x = days_since_prior_order)) +
  geom_bar(fill = c(rep("grey25", 6), "orange", "orange",  rep("grey25", 5), "orange",
                    rep("grey25", 6), "orange", rep("grey25", 6), "orange",
                    rep("grey25", 3))) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Day 6",
       x = "Day Since Last Order")

grid.arrange(po0, po1, po2, po3, po4, po5, po6,
           ncol = 1)
```

### Order Hour of Day when orders on the same day {.tabset}

There are a sizeable number of orders that happen on the same day, represented
by days_since_prior_order == 0. This raises some questions:

- When are they making the second order for the day?
- How many hours apart are they making orders if ordered on the same day?

#### When is the second order?

```{r}
# Select orders where days_since_prior_order == 0
ggplot(orders[as.numeric(orders$days_since_prior_order) == 0, ], 
       aes(x = order_hour_of_day)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(x = "Hour of Day",
       y = "",
       title = "Hour of day when days since prior order = 0")
```

#### How far apart?

If someone is ordering on the same day, they are mostly ordering within the same
hour, or shortly after their first order. Good thing they noticed soon after
their original order!

```{r}
## Calculate hours since prior order, only if days since prior order is 0
orders <- data.table(orders)
temp <- orders$order_hour_of_day
temp <- c("24", temp)
temp <- temp[1:length(temp)-1]
orders$temp <- temp
orders[, hours_since_last_order := as.numeric(order_hour_of_day) - as.numeric(temp)]
orders[days_since_prior_order != 0 | is.na(days_since_prior_order), 
       hours_since_last_order := NA]
temp <- orders[days_since_prior_order == 0, ]

ggplot(temp, aes(x = hours_since_last_order)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(x = "Hours Since Last Order",
       y = "",
       title = "Hours Since Last Order if Ordered on the Same Day")
```

## Orders by Day of the Week

There are a lot of orders on each day of the week. However, there are 
significantly more orders on days 0 and 1. 

Could this represent the weekend?

```{r orders_day_of_week}

ggplot(orders, aes(x = order_dow)) + 
  geom_bar(fill = c(rep("gold", 2), rep("grey25", 5))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(x = "Day of Week (unknown)",
       y = "Count",
       title = "Distribution of Orders by Day of Week") +
  scale_y_continuous(labels = comma)

```


## Orders by Hour of the Day

Here we can see at what hours of the day people are making orders.

Reading the plot from left to right:

- There are few orders between 00 - 05, representing few orders during sleeping hours
- Orders pick up beginnig at hour 06 as people start to wake up
- Orders continue to pick up rapidly in the morning, peaking at hour 10 (gold)
- Orders continue to be made during the morning, experiencing a small dip at noon
- There is a slight pick up of orders just after lunch at hour 15
- Orders begin to drop commencing at hour 16 through to hour 23

```{r orders_hour_of_day}
ggplot(orders, aes(x = order_hour_of_day)) + 
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 4), "cyan3",
                    rep("grey25", 8))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(x = "Hour of the Day",
       y = "",
       title = "Distribution of Orders by Hour of the Day") +
  scale_y_continuous(labels = comma)

```

So most orders are made between the hours 09 and 16. But does this pattern exist
over all days of the week?

### Order by Hour and Day of Week

In the graph below, the pattern we observed is no longer observed.

Instead, Day 1 to Day 5 has hour 10 as the most popular hour for orders, while
days 0 and 6 have hour 14 as the most popular. It suggests to me that Days 1 - 5
are weekdays, while Days 0 and 6 are weekends.

```{r hour_and_dow, fig.width = 7, fig.height = 8}
# ggplot(orders, aes(x = order_hour_of_day)) +
#   geom_bar() +
#   theme_minimal() +
#   theme(axis.ticks.x = element_blank(),
#         axis.ticks.y = element_blank(),
#         axis.text.y = element_blank(),
#         legend.position = "none",
#         panel.grid.major = element_blank()) +
#   labs(x = "Hour of Day",
#        y = "",
#        title = "Distribution of Orders by Hour of Day") +
#   facet_grid(order_dow ~ .)

p0 <- ggplot(orders[order_dow == 0, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 14), "gold", rep("grey25", 9))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 0")

p1 <- ggplot(orders[order_dow == 1, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 13))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 1")

p2 <- ggplot(orders[order_dow == 2, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 13))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 2")

p3 <- ggplot(orders[order_dow == 3, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 13))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 3")

p4 <- ggplot(orders[order_dow == 4, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 13))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 4")

p5 <- ggplot(orders[order_dow == 5, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 10), "gold", rep("grey25", 13))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 5")

p6 <- ggplot(orders[order_dow == 6, ], aes(x = order_hour_of_day)) +
  geom_bar(fill = c(rep("grey25", 14), "gold", rep("grey25", 9))) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Day 6",
       x = "Hour of the Day")



grid.arrange(p0, p1, p2, p3, p4, p5, p6, ncol = 1)

```